---
title: "2. Audit-class case studies — LGALS3BP / LTBP1 / ALB / GAPDH"
author: "Jibeom Ko"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{2. Audit-class case studies (4 genes, one per audit class)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.0,
  fig.height = 3.5,
  dpi = 110)
```

# Goal

Walk through **one representative gene per audit_class** to show how
the v0.3.0 audit framework handles different evidence patterns. None
of these genes are presented as the paper's flagship candidate; each
is an *illustrative case* of how the framework's three axes plus two
gates resolve a specific evidence pattern into a transparent label.

| Gene | audit_class | Why it ends here |
|---|---|---|
| **LGALS3BP** | `high_confidence` | Both gates clean; multi-layer + Early-onset evidence + serum bridge all strong; recovered as a secondary anchor |
| **LTBP1** | `supported_uncertain` | Multi-layer concordant tissue evidence, but cohort I² = 75% triggers `heterogeneity_gate < 1`; Class B inverse-translation exemplar |
| **ALB** | `penalized` | Plasma high-abundance flag triggers `leakage_gate = 0.5`; tissue Early signal is real but not specific enough on its own |
| **GAPDH** | `excluded` | Housekeeping flag triggers `leakage_gate = 0`; hard-zeros the score regardless of trajectory shape |

```{r setup}
library(pdactrace)
library(data.table)
```

# Side-by-side audit summary

```{r side-by-side}
case_genes <- c("LGALS3BP", "LTBP1", "ALB", "GAPDH")
audit <- compute_audit_score(case_genes)
audit
```

Two key observations:

1. **`audit_class` is determined by gates first**, then by score. GAPDH
   (housekeeping → `leakage_gate = 0`) is `excluded` regardless of how
   strong its trajectory is. ALB (plasma high-abundance →
   `leakage_gate = 0.5`) is `penalized` even though its tissue signal
   is real.
2. **LTBP1 is *supported_uncertain*, not `high_confidence`**, because
   the heterogeneity gate (`audit_heterogeneity_gate = 0.7`) marks
   cohort divergence; we frame it as an exemplar of Class B inverse
   translation rather than a flagship candidate.

# Detailed audit features per gene

```{r features}
detailed <- query_panel(case_genes)
extra <- pdactrace_reference[
  gene_symbol %in% case_genes,
  .(gene_symbol, audit_class, audit_score, max_I2_meta)]
detailed <- merge(detailed, extra, by = "gene_symbol", sort = FALSE)
detailed[, .(gene_symbol, rna_pattern, prot_pattern, translation_class,
              audit_class, audit_score, max_I2_meta,
              cell_origin_top)]
```

# Case 1 — LGALS3BP (`high_confidence` anchor)

LGALS3BP (Galectin-3 binding protein) sits at the top of the
`audit_score` ranking and is one of the secondary anchors recovered
by the framework. It illustrates the simplest case: every axis fires
positively and neither gate triggers.

```{r lgals3bp-detail}
lg <- query_gene_detailed("LGALS3BP")
lg$summary
lg$per_stage
```

```{r lgals3bp-plot, fig.width = 4.5, fig.height = 3.5}
plot_gene_hexagon("LGALS3BP",
                    comparison = "high_confidence_mean")
```

# Case 2 — LTBP1 (`supported_uncertain`, Class B inverse exemplar)

LTBP1 has tissue-level concordant evidence (RNA + protein both
`Early_Burst_Up`, ρ ≈ 1.0, myCAF cell origin) but cohort-level
heterogeneity is high: TCGA shows an increasing trajectory, CPTAC a
decreasing trajectory, and two cohorts are non-monotonic. The
random-effects meta-analysis records `max_I2_meta = 75 %` and the
heterogeneity gate downweights the score by 0.70.

```{r ltbp1-detail}
ltbp1 <- query_gene_detailed("LTBP1")
ltbp1$per_cohort
```

The framework correctly assigns LTBP1 to `supported_uncertain` rather
than `high_confidence`, traceably attributing the uncertainty to
cohort divergence:

```{r ltbp1-plot, fig.width = 4.5, fig.height = 3.5}
plot_gene_hexagon("LTBP1",
                    comparison = "high_confidence_mean")
```

LTBP1 also has `translation_class = "B"` (inverse): tissue
Burst-Up evidence does *not* translate to a serum-up biomarker.
Its panel value emerges only when paired with a Class A
(concordant-translation) marker — see *Class A × Class B panel
diversity* in the BiB Discussion (LTBP1 + SERPINA1 LOOCV AUC 0.97
vs LTBP1 alone 0.83 vs SERPINA1 alone 0.89).

# Case 3 — ALB (`penalized`, plasma high-abundance gate)

Albumin (ALB) has a real Early-onset signal at the tissue level
(`Early_Trough` — falling at PDAC onset) but is also one of the most
abundant plasma proteins. The leakage gate downweights ALB's score
by 0.50 to reflect that abundance-driven false positives are a known
failure mode in serum biomarker discovery.

```{r alb-detail}
alb <- compute_audit_score("ALB")
alb[, .(gene_symbol, evidence_strength, biological_coherence,
         translational_relevance, leakage_gate,
         heterogeneity_gate, audit_score, audit_class)]
```

```{r alb-plot, fig.width = 4.5, fig.height = 3.5}
plot_gene_hexagon("ALB", comparison = "penalized_mean")
```

ALB is admitted to the ranking (its real Early signal counts) but
with a transparent abundance-driven discount, rather than being
silently dropped. This is the difference between a "leakage gate"
(graded 0 / 0.5 / 1) and a hard exclusion list.

# Case 4 — GAPDH (`excluded`, housekeeping gate)

GAPDH is a canonical housekeeping gene. Its `audit_is_housekeeping`
flag triggers `leakage_gate = 0`, which hard-zeros the audit score
regardless of how strong its per-stage trajectory looks. This is the
"non-nominatable" failure mode of the framework: even a perfect
trajectory shape cannot rescue a housekeeping artifact.

```{r gapdh-detail}
gapdh <- compute_audit_score("GAPDH")
hk <- pdactrace_reference[gene_symbol == "GAPDH",
                            .(gene_symbol, audit_is_housekeeping)]
gapdh <- merge(gapdh, hk, by = "gene_symbol", sort = FALSE)
gapdh[, .(gene_symbol, audit_is_housekeeping, leakage_gate,
           audit_score, audit_class)]
```

```{r gapdh-plot, fig.width = 4.5, fig.height = 3.5}
plot_gene_hexagon("GAPDH", comparison = "excluded_mean")
```

GAPDH is also reclassified under v0.4.0's 12-template catalog as
`Monotonic_Up` (linear pan-stage progression) rather than
`Early_Burst_Up`, which removes it from the Early × 4 atlas surface
entirely — the leakage gate and the 12-template catalog discipline
defend against housekeeping contamination through two independent
mechanisms.

# Take-away

The four cases illustrate the **explicit decomposition** that the
audit framework provides: each gene's `audit_class` is traceable to
specific axes and gates, and the same framework discipline applies
uniformly across genes regardless of biological role. When you query
your own candidate, the same decomposition is available via
`compute_audit_score(c("YOUR_GENE"))` and `query_gene_detailed(...)` —
see vignette `user_cohort_extension` for applying this discipline
to your own multi-omics cohort.
