---
title: "3. Audit framework: 3 axes + 2 gates + 4 classes (v0.3.0)"
author: "Jibeom Ko"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Audit framework: 3 axes + 2 gates + 4 classes (v0.3.0)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r audit-framework-01, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.0,
  fig.height = 4.0,
  dpi = 110)
```

# Goal

This vignette walks through the v0.3.0 audit framework — a transparent
evidence aggregation layer for PDAC early-marker hypothesis
prioritization. The framework summarises each gene along **three
evidence axes** (`evidence_strength`, `biological_coherence`,
`translational_relevance`), applies **two reliability gates**
(`leakage_gate`, `heterogeneity_gate`), and assigns it to one of
**four audit classes** (`high_confidence`, `supported_uncertain`,
`penalized`, `excluded`). Unlike a supervised biomarker predictor,
the rule combines pre-specified weights with deterministic gating
and Monte Carlo uncertainty propagation, exposing cohort
heterogeneity, layer agreement, and artifact risk.

```{r audit-framework-02, message = FALSE, warning = FALSE}
library(pdactrace)
library(data.table)
```

# 1. Why a hand-engineered rule (and not a supervised model)

PDAC early-stage diagnosis lacks a validated ground-truth biomarker
beyond CA19-9 (Mehta *et al.*, BJS Open 2024 prediagnostic
meta-analysis). Training a supervised classifier on community-curated
markers risks two well-known failure modes:

1. **Internal circularity** — using `pdactrace`'s own pipeline outputs
   (e.g., the phase29 resectable list) as labels would teach the model
   to reproduce its own heuristic.
2. **Anchor circularity** — using literature-curated anchors as
   training labels collapses the tool into a "literature mimic": it
   learns to prioritize what other papers already prioritized, with no
   genuine discovery capability.

The audit framework therefore declines to predict and instead
**aggregates, ranks, and reports uncertainty**. External anchors are
reserved for post-freeze enrichment evaluation only.

# 2. The 6-function API at a glance

```{r audit-framework-03}
# 1. Build a per-gene evidence graph (nodes = layers / multipliers / readouts)
g <- build_evidence_graph("LTBP1")
g$nodes

# 2. Extract the deterministic feature components (used internally to
#    compose the three axes; mapping table is in supplement S1)
extract_graph_features("LTBP1")

# 3. Compute the 3-axis + 2-gate audit score and 4-class label
compute_audit_score("LTBP1")

# 4. Propagate uncertainty by Monte Carlo (N=200 here for speed)
propagate_uncertainty("LTBP1", n_mc = 200, seed = 42)

# 5. Evaluate anchor enrichment in the top-N audit-ranked genes
evaluate_anchor_enrichment(top_n = c(50, 100, 500), tier = "secondary")

# 6. Curate a transparent case-study label table
case_study(c("LTBP1", "GAPDH", "THBS2", "ALB", "AMBP", "SERPINA1"))
```

# 3. LTBP1 — the `supported_uncertain` exemplar

LTBP1 illustrates the framework's most distinctive behaviour: a gene
whose pooled-RNA signal is `Early_Burst_Up` (rho ≈ 1.0) but whose
per-cohort meta-analysis exposes severe heterogeneity (max I² ≈ 75 %,
TCGA up vs CPTAC down). The frozen rule down-weights LTBP1 via the
heterogeneity gate (0.70 in the 70-90 band), and assigns it to the
`supported_uncertain` audit class — protein, scRNA, and serum layers
all raise the `evidence_strength` axis, but the heterogeneity gate
prevents a high-confidence call.

```{r audit-framework-04}
ltbp1_score <- compute_audit_score("LTBP1")
ltbp1_score
ltbp1_unc <- propagate_uncertainty("LTBP1", n_mc = 200, seed = 42)
ltbp1_unc
```

The wide 95% CI (roughly `[0.26, 0.88]` at N=500 MC iterations) is the
honest readout: LTBP1 is a credible candidate but its uncertainty is
real and traceable to cohort divergence, not buried under a single
point estimate.

# 4. GAPDH — the housekeeping negative control

GAPDH carries the strict housekeeping flag and therefore receives a
zero leakage multiplier. The score is exactly 0 in every Monte Carlo
iteration, and the gene is classified as `excluded`.

```{r audit-framework-05}
compute_audit_score("GAPDH")
propagate_uncertainty("GAPDH", n_mc = 200, seed = 42)
```

This is the deliberate behaviour: a transparent audit framework should
refuse to nominate housekeeping genes regardless of how well their
trajectory shape fits an early-onset template.

# 5. Anchor enrichment evaluation (post-freeze)

External literature anchors are organised into three provenance tiers
(`primary`, `secondary`, `exploratory`). Enrichment is computed with
the audit_score frozen — anchors never participate in scoring.

```{r audit-framework-06}
evaluate_anchor_enrichment(top_n = c(50, 100, 200, 500),
                           tier = c("primary", "secondary"))
```

A statistically significant enrichment (p ≪ 0.001) of T1+T2 anchors in
the top 100 indicates that the deterministic rule recovers
community-prioritized hypotheses without ever having seen them.

# 6. Reading the `case_study()` output

```{r audit-framework-07}
case_study(c("LTBP1", "GAPDH", "THBS2", "ALB", "AMBP",
             "SERPINA1", "ANPEP"))
```

Each gene receives an `audit_class` (4-class output of the frozen
rule) and a `pdactrace_call` (provenance-aware label that combines
class with anchor / tier / artifact context):

- **audit_class**
  - `high_confidence` — gates clean and audit_score ≥ 0.5
    (e.g., LGALS3BP, LCN2, GGH, THBS2).
  - `supported_uncertain` — heterogeneity-gated or score in the
    0.3-0.5 band (e.g., LTBP1, FN1, TGFBI, SLC16A3).
  - `penalized` — plasma high-abundance (leakage_gate = 0.5;
    e.g., ALB, FGB, C6).
  - `excluded` — housekeeping (leakage_gate = 0; e.g., GAPDH).
- **pdactrace_call**
  - `confirmed_anchor` — externally validated marker that ranks high
    (e.g., THBS2, TIMP1, MUC16, CEACAM1).
  - `rescued_candidate` — RNA-tier ARTIFACT/OTHER but multi-layer
    supported (e.g., LTBP1, VCAN, SPARC).
  - `novel_nomination` — non-anchor candidate with strong multi-layer
    signal and a clean leakage gate (e.g., LGALS3BP, KLK7/10, CD55).
  - `rejected_artifact` — housekeeping or plasma high-abundance
    (e.g., GAPDH, ALB, C6) — surfaced only for negative-control display.

Together with the Monte Carlo `confidence_class` (`stable_high`,
`high_uncertain`, `medium`, `low`, `excluded`) this gives a
transparent two-axis call: *which class does the deterministic rule
assign* and *how stable is the score under realistic perturbation*.

# 7. What the audit framework does not claim

To stay defensible, the framework explicitly does not:

- predict a probability of being a "true PDAC early marker"
  (no ground truth exists);
- learn weights from anchors or from `pdactrace`'s own internal
  pipeline outputs;
- collapse high-heterogeneity signals into a single point estimate
  without a confidence interval;
- treat housekeeping or plasma high-abundance proteins as candidates,
  even when their trajectory shape is suggestive.

These are deliberate design choices recorded in the supplement and the
`NEWS.md` v0.3.0 entry.

# Session info

```{r audit-framework-08}
sessionInfo()
```
