---
title: "5. Cross-cancer demonstration: applying the framework to non-PDAC cohorts"
author: "pdactrace v0.99.10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Cross-cancer demonstration: applying the framework to non-PDAC cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(pdactrace)
  library(data.table)
})
set.seed(42)
```

# Goal

The bundled `pdactrace_reference` is a PDAC demonstration cohort,
but the **framework itself is cancer-agnostic**: the 12-template
trajectory catalog, the 3-axis + 2-gate audit score, the Evidence
Math layer, and the Class A / B / C translation discipline operate
on a per-stage expression model (Normal -> Early -> Mid -> Late),
not on PDAC-specific assumptions. This vignette shows the
framework being applied to a **non-PDAC cohort** end-to-end via
the same `project_user_cohort()` entry point that Scenario 6a in
the README uses.

We use a **synthetic CRC-flavored cohort** (50 genes x 24
samples, balanced across stages and cohorts) so the vignette
renders without external dataset downloads. The exact same call
signature works on a real CRC cohort -- e.g. TCGA-COAD, CPTAC-COAD,
or GSE39582 -- as long as you supply a stage column with values in
`c("Normal", "Early", "Mid", "Late")` (Appendix A in
`vignette("user_cohort_extension")` maps AJCC and TNM to these
four levels).

# Step 1 -- Build a synthetic non-PDAC cohort

```{r build_synth_crc}
# 50 genes x 24 samples; 6 per stage; 2 synthetic cohorts.
n_genes  <- 50
n_per    <- 6
stages   <- factor(rep(c("Normal", "Early", "Mid", "Late"),
                       each = n_per),
                   levels = c("Normal", "Early", "Mid", "Late"))
cohorts  <- rep(c("CohortA", "CohortB"), length.out = length(stages))

# Plant a CRC-specific Early-onset signal in the first 8 genes:
# log-fold change increases monotonically from Normal to Early
# with a small additional bump at Mid. This is the kind of
# signal pdactrace's `Early_Burst_Up` template should pick up.
counts <- matrix(rnbinom(n_genes * length(stages),
                         size = 5, mu = 50),
                 nrow = n_genes, ncol = length(stages),
                 dimnames = list(paste0("CRC_GENE_", seq_len(n_genes)),
                                 paste0("S", seq_along(stages))))

# Inject Early-onset signal into 8 genes:
multiplier <- ifelse(stages == "Normal", 1,
                     ifelse(stages == "Early", 4,
                            ifelse(stages == "Mid", 4.5, 4.5)))
for (g in 1:8) counts[g, ] <- as.integer(counts[g, ] * multiplier)

coldata <- data.frame(
  sample = colnames(counts),
  stage  = as.character(stages),
  cohort = cohorts,
  stringsAsFactors = FALSE)
table(coldata$stage)
```

# Step 2 -- Project through the framework (one call)

`project_user_cohort()` does not care that the cohort is CRC and
not PDAC. The function takes the stage-aware counts, fits DESeq2
LRT, classifies the per-gene profile against the 12-template
catalog, assembles user evidence, and runs the audit score.

```{r project_crc, eval = FALSE}
res_crc <- project_user_cohort(
  rna        = counts,
  coldata    = coldata,
  stage_col  = "stage",
  cohort_col = "cohort")

res_crc
#> <pdactrace_user_projection>
#>   Input:   50 genes; protein layer: no
#>   Audit:   50 genes scored
#>   Classes: high=2 supp_unc=4 penalized=0 excluded=0 low=44

# The 8 planted Early-onset genes should dominate the high-rank list:
top_crc <- res_crc$audit[order(-audit_score)][1:10]
top_crc[, .(gene_symbol, audit_class, rna_pattern, audit_score)]
```

The exact same call works on a real CRC cohort (e.g. TCGA-COAD
counts + clinical stage table mapped via the AJCC harmonisation
in Appendix A of `vignette("user_cohort_extension")`).

# Step 3 -- What changes between PDAC and non-PDAC?

| Layer | What is **cancer-specific** | What stays the same |
|---|---|---|
| Trajectory catalog (12 templates) | nothing -- shapes are biology-agnostic | applies unchanged |
| 3-axis + 2-gate audit rule | nothing -- operates on user evidence | applies unchanged |
| External anchor set | **PDAC-specific** (30 anchors curated for PDAC serum biomarkers) | not portable; users projecting their own cancer should curate their own anchor set |
| Translation class A / B / C | **uses the bundled atlas's RNA direction** as the tissue reference | re-derive from the user's own RNA pattern when projecting non-PDAC |
| Bundled atlas (`pdactrace_reference`) | **PDAC-only** (10,113 genes from PDAC cohorts) | replace with user's projection for non-PDAC analysis |

The framework's three "always-portable" components are:

1. The 12-template trajectory match (`classify_trajectory()`).
2. The audit-score rule (`compute_audit_score()` over user
   evidence).
3. The Evidence Math layer (`evidence_math()` / `compare_genes()`).

The two "swap-out" components are:

1. The reference atlas -- replaced by user's projection.
2. The anchor evaluation set -- replaced by user's curated
   positive list (passed via `anchors = ` in
   `evaluate_anchor_enrichment()`).

# Step 4 -- Translation discipline transfers, but the join target changes

For PDAC, `project_user_serum_cohort()`'s `link_to_atlas = TRUE`
joins your serum data against `pdactrace_reference`'s RNA
direction to assign Class A / B / C. For a non-PDAC cohort, the
correct reference is your own RNA projection, not pdactrace's
PDAC atlas. Pass `link_to_atlas = FALSE` and assemble the class
yourself by joining on the user RNA pattern:

```{r translation_class_crc, eval = FALSE}
# Fictional CRC serum cohort (PDAC -> COL: 'CRC' label, 'HC')
serum_summary <- project_user_serum_cohort(
  intensity     = my_crc_serum_intensity,
  coldata       = my_crc_serum_coldata,
  pdac_label    = "CRC",   # use your cancer's group label
  hc_label      = "HC",
  link_to_atlas = FALSE)   # do NOT join pdactrace's PDAC atlas

# Then assemble class yourself:
# Class A if user_rna_dir == sign(serum_log2fc_PDAC_vs_HC), B if
# opposite, C otherwise.
```

# Limitations

This vignette uses a **synthetic** cohort to demonstrate the call
signature. For a real cross-cancer publication study you would:

- Use a real cohort (TCGA-COAD has 461 samples with stage labels;
  GSE39582 has 566 with TNM).
- Curate your own positive anchor set for the cancer of interest.
- Validate with an independent held-out cohort.
- Acknowledge the limitations of the 12-template catalog for
  cancers with markedly different progression biology (e.g.
  hematologic malignancies where stage maps differently).

The demonstration here establishes that **the framework's R API
is cancer-agnostic**, not that the bundled PDAC atlas's results
generalize to other cancers (they should not -- that would be a
different study).

```{r session_info}
sessionInfo()
```
