---
title: "1. Lookup basics: querying the pdactrace atlas"
author: "Jibeom Ko"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Lookup basics: querying the pdactrace atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 3.0,
  dpi = 110)
```

# Goal

`pdactrace` is a queryable PDAC multi-omics gene-level lookup atlas.
This vignette walks through the **5 core lookup functions** in the
order most users will need them.

```{r setup}
library(pdactrace)
```

# 1. Single-gene query — `query_gene()`

```{r}
q <- query_gene("LTBP1")
print(q)
```

Each layer is accessible as a named slot:

```{r}
q$rna
q$serum
q$filter_status
```

# 2. Multi-gene comparison — `query_panel()`

```{r}
qp <- query_panel(c("LTBP1", "SERPINA1", "CDH13", "SPARC", "CP", "FGB"))
qp[, .(gene_symbol, rna_pattern, translation_class,
        flt_signal_peptide, flt_final, panel_member)]
```

Genes not in the LRT-significant atlas universe are reported and
skipped:

```{r}
qp_miss <- query_panel(c("LTBP1", "FOOBAR_NOT_A_GENE"))
```

# 3. Multi-parameter candidate listing — `list_candidates()`

The new v0.1.0 signature accepts biology-level filters. Pass `NULL`
(default) for any parameter to leave it unconstrained.

```{r}
# Class B (inverse stromal) candidates
list_candidates(translation_class = "inverse")[
  , .(gene_symbol, rna_pattern, translation_class, panel_member)]

# Compound filter: signal-peptide-positive panel members
list_candidates(signal_peptide = TRUE, panel_member = TRUE)[
  , .(gene_symbol, rna_pattern, translation_class, panel_member)]

# Direction-aware: tissue-Up + serum-detected
nrow(list_candidates(tissue_direction = "Up", serum_detected = TRUE))
```

# 4. Filter audit trail — `trace_filters()`

Shows which `phase60` 7-step filters each gene passed plus auxiliary
route flags (`phase77_strict`, `panel_member`, `resectable_marker`).
Most genes pass *one* of the two narrowing funnels but not both —
this is the audit message of pdactrace.

```{r}
trace_filters(c("SERPINA1", "LTBP1", "SPARC", "CDH13"))[
  , .(gene_symbol, n_phase60_pass, class_route)]
```

`SERPINA1` passes all 7 (`phase60_final`); `LTBP1`, `SPARC`, `CDH13`
each pass only `SignalP` but are surfaced via `phase77_classB`. Both
routes are valid — the tool shows them honestly.

# 5. Plain-text summary — `summarize_gene_evidence()`

```{r}
cat(summarize_gene_evidence("LTBP1"))
```

# Atlas metadata

```{r}
str(list_atlas_metadata(), max.level = 1)
```

# Next

See vignette `02_ltbp1_case_study` for the full LTBP1 narrative
including visualizations (`plot_gene_evidence`, `plot_filter_trace`,
`plot_panel_heatmap`, `plot_candidate_landscape`).

# Session

```{r}
sessionInfo()
```
