---
title: "6. Methodology validation: cutoff sensitivity + tool comparison"
author: "pdactrace v0.99.10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Methodology validation: cutoff sensitivity + tool comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(pdactrace)
  library(data.table)
})
```

# Goal

Two reviewer-facing methodology **sanity checks** on the bundled
`pdactrace_reference` atlas (post-freeze evaluation, not
fully-blinded external validation):

- **Section A -- Cutoff sensitivity.** How robust is the framework's
  top-N anchor enrichment to perturbations of the trajectory
  classification threshold (`rho_cutoff`) and the audit-score
  axis weights?
- **Section B -- Tool comparison.** Does the audit_score-based
  ranking actually beat simple alternatives -- raw |beta| ranking,
  -log10(padj) ranking, or signal-peptide-only filtering --
  measured on the same external anchor set?

Both sections run on the bundled atlas; no external download or
per-stage refit required.

```{r load_atlas}
ref <- pdactrace:::.get_reference()
cat("Atlas dimensions:", dim(ref), "\n")
data("pdactrace_external_anchors", package = "pdactrace")
t1_genes <- pdactrace_external_anchors$gene[
  pdactrace_external_anchors$evidence_tier == "T1_validated"]
cat("T1-validated anchors in atlas:",
    sum(t1_genes %in% ref$gene_symbol), "/", length(t1_genes),
    "\n")
```

# Section A -- Cutoff sensitivity

## A1. Trajectory classification cutoff (`rho_cutoff`)

The default `rho_cutoff = 0.85` means a gene's z-scored 4-stage
profile must correlate with one of the 12 templates at Pearson
`rho >= 0.85` to be called as that pattern. Below the threshold,
the gene goes into `rna_pattern = NA` (no surface call).

```{r rho_sweep}
rho_grid <- c(0.80, 0.85, 0.90)
rho_sweep <- rbindlist(lapply(rho_grid, function(co) {
  # Re-classify by simulating a stricter / looser cutoff on the
  # bundled `rna_pattern_rho` (which carries the actual best-match
  # rho per gene). Surface call survives only if rho >= cutoff.
  surface_kept <- ref[!is.na(rna_pattern_rho) &
                       rna_pattern_rho >= co, gene_symbol]
  data.table(
    rho_cutoff   = co,
    n_surfaced   = length(surface_kept),
    pct_surfaced = round(100 * length(surface_kept) /
                          nrow(ref), 1))
}))
print(rho_sweep)
```

**Interpretation.** Around 89% of atlas genes meet the default
`rho >= 0.85` threshold; loosening to `0.80` adds ~3 percentage
points (mostly noisy edge cases) and tightening to `0.90` drops
~4 points (cuts moderate-fit genes). The atlas's surface decisions
overlap heavily across the `[0.80, 0.90]` range -- the framework
is robust to modest perturbations. The bundled `audit_score`'s
top-100 anchor enrichment (`39.3x`, `p = 2.2e-10`) is computed at
the `rho = 0.85` surface and would be recomputed if a user re-ran
the full pipeline at a different cutoff via
[classify_trajectory()].

## A2. Three-axis weight sensitivity

The default weights are `0.40 / 0.35 / 0.25` for
`evidence_strength / biological_coherence /
translational_relevance`. We sweep across the full simplex --
including single-axis extremes -- and recompute `audit_score`
for the bundled atlas, then re-evaluate top-100 anchor enrichment.

```{r weight_sweep}
# Pull the three sub-component scores already in the atlas.
sub <- ref[, .(gene_symbol,
               ev_strength = audit_evidence_strength,
               bio_coh     = audit_biological_coherence,
               trans_rel   = audit_translational_relevance,
               leak_mult   = audit_leakage_mult,
               het_mult    = audit_heterogeneity_mult)]

weight_grid <- list(
  default       = c(0.40, 0.35, 0.25),
  ev_only       = c(1.00, 0.00, 0.00),  # extreme: only layer presence
  bio_only      = c(0.00, 1.00, 0.00),  # extreme: only biology
  trans_only    = c(0.00, 0.00, 1.00),  # extreme: only serum reach
  inverted      = c(0.25, 0.35, 0.40))  # default reversed

weight_eval <- rbindlist(lapply(names(weight_grid), function(nm) {
  w <- weight_grid[[nm]]
  pos <- w[1] * sub$ev_strength + w[2] * sub$bio_coh +
         w[3] * sub$trans_rel
  pos[!is.finite(pos)] <- 0
  scored <- copy(ref)
  scored[, audit_score := pos * sub$leak_mult * sub$het_mult]
  enr <- evaluate_anchor_enrichment(top_n = 100, tier = "secondary",
                                     evidence = scored)
  data.table(profile = nm,
             weights = paste(sprintf("%.2f", w), collapse = "/"),
             hits    = enr$hits,
             fold    = round(enr$fold, 1),
             pval    = formatC(enr$pval, format = "g", digits = 2))
}))
print(weight_eval)
```

**Interpretation.** Top-100 anchor enrichment is identical
(`fold = 39.3`, `p = 2.2e-10`) across the **entire weight
simplex** -- including single-axis extremes (1.00 / 0 / 0) and
the default-inverted profile (0.25 / 0.35 / 0.40). The framework
is therefore **fundamentally insensitive to the precise weight
choice** within the simplex. Two reasons drive this robustness:

1. **Multiplicative gates dominate the ranking.** The leakage
   gate (multiplier in `{0.0, 0.5, 1.0}`) and heterogeneity gate
   (multiplier in `{0.3, 0.7, 1.0}`) collapse a large fraction
   of the atlas to small or zero scores; the top-100 set is
   determined more by which genes survive both gates than by
   how the surviving gene's three axis scores are weighted.
2. **The three axes are positively correlated** in the bundled
   atlas (a gene with strong layer presence tends to have strong
   biological coherence and translational reach). Any weighted
   sum therefore produces a similar ranking.

This is a **strong defense against the "why those exact weights?"
reviewer question**: the answer is that the framework's
robustness does not come from precise weight tuning, but from
the gate structure plus axis correlation in the underlying
biology.

# Section B -- Tool comparison

How does the full audit-score ranking compare to simpler
alternatives that a reviewer might propose? We evaluate each
ranking on the same anchor harness.

```{r tool_compare}
# Build candidate ranking columns on the bundled atlas:
ranking_cols <- list(
  audit_score   = ref$audit_score,
  abs_beta_meta = ref$max_abs_beta_meta,
  neglog10_padj = -log10(pmax(ref$rna_lrt_padj,
                               .Machine$double.eps)),
  layer_count   = (!is.na(ref$rna_pattern))     +
                  (!is.na(ref$prot_pattern))    +
                  (!is.na(ref$cell_origin_top)) +
                  ifelse(is.na(ref$serum_detected), FALSE,
                          ref$serum_detected))

# Hypergeometric anchor enrichment over the top-N of an arbitrary
# ranking column. Replicates the math used inside
# evaluate_anchor_enrichment() but accepts any ranking vector.
.hypergeo_topn <- function(rank_vec, anchor_set, top_n) {
  n_atlas <- length(rank_vec)
  n_in    <- sum(anchor_set %in% ref$gene_symbol)
  ord     <- order(-rank_vec, na.last = TRUE)
  top_g   <- ref$gene_symbol[ord[seq_len(top_n)]]
  hits    <- sum(top_g %in% anchor_set)
  expect  <- top_n * n_in / n_atlas
  pval    <- stats::phyper(hits - 1, n_in, n_atlas - n_in,
                            top_n, lower.tail = FALSE)
  data.table(top_n = top_n, hits = hits,
             expect = round(expect, 2),
             fold = round(hits / max(expect, 1e-9), 1),
             pval = formatC(pval, format = "g", digits = 2))
}

# Secondary-tier anchor set (T1 + T2):
sec_anchors <- pdactrace_external_anchors$gene[
  pdactrace_external_anchors$include_secondary_eval]

rank_eval <- rbindlist(lapply(names(ranking_cols), function(nm) {
  out <- rbindlist(lapply(c(50, 100, 200), function(N) {
    .hypergeo_topn(ranking_cols[[nm]], sec_anchors, N)
  }))
  out[, ranking := nm]
  out
}))
print(rank_eval[, .(ranking, top_n, hits, expect, fold, pval)])
```

**Interpretation.** Head-to-head on the same hypergeometric test
against the same secondary-tier anchor set, the four rankings
produce dramatically different top-N performance:

| Ranking | top-50 fold | top-100 fold | top-200 fold | Reads |
|---|---|---|---|---|
| `audit_score` | **44.9** | **39.3** | 22.5 | best at the precision-critical top |
| `layer_count` | 33.7 | 28.1 | **28.1** | best at top-200 |
| `neglog10_padj` | 11.2 | 5.6 | 2.8 | weak; biased toward well-powered genes |
| `abs_beta_meta` | 0.0 | 0.0 | 2.8 | fails; ranks plasma-abundant artifacts high |

The two simplest rankings (`abs_beta_meta`, `neglog10_padj`)
fail badly. **The full `audit_score` dominates at the most
precision-critical top-50 and top-100** -- by both hit count
and fold-enrichment -- which is the regime that matters for a
candidate prioritisation tool that you actually pull a panel
from. `layer_count` (the next-best simpler alternative,
combining multi-layer evidence presence) catches up in raw hits
at top-200, but loses the early-rank specificity advantage that
the gates provide.

This is the head-to-head comparison reviewers ask for: "does
your score actually do something a simpler rank does not?" The
quantitative answer is **yes at the precision-critical top of
the list**, where biomarker prioritisation actually matters.

# Section C -- Negative control depletion

A complementary sanity check: if the anchor enrichment in the
top-N tells us "the framework correctly elevates known
biomarkers", the **negative control depletion** in the
top-N (or equivalently, **enrichment in the bottom-N**) tells
us "the framework correctly suppresses known artefacts". The
two together demonstrate that the framework discriminates in
both directions, not only the positive direction. This is the
direct test reviewers ask for when they suspect an enrichment
result might be circular: if your negative controls were also
enriched, you'd worry about an aspecific signal; if they're
depleted, the framework is doing real work.

We use the two negative-control flags already in the bundled
atlas: `audit_is_housekeeping` (n = 31; canonical housekeeping
genes whose stage-trajectory signal is presumed artifactual)
and `audit_is_plasma_high_abundance` (n = 22; top-decile plasma
proteins like ALB / IGHG1 elevated in any inflammatory state).
Both are excluded from the audit-score's positive scoring via
the **leakage gate** (housekeeping multiplier `0.00`, plasma_high
multiplier `0.50`).

```{r negctrl_setup}
hk_set  <- ref$gene_symbol[!is.na(ref$audit_is_housekeeping) &
                            ref$audit_is_housekeeping]
plh_set <- ref$gene_symbol[!is.na(ref$audit_is_plasma_high_abundance) &
                            ref$audit_is_plasma_high_abundance]
cat("Negative control sets:\n",
    "  housekeeping:        ", length(hk_set), " genes\n",
    "  plasma_high_abundance:", length(plh_set), " genes\n", sep = "")

n_atlas <- nrow(ref)
ord_desc <- order(-ref$audit_score, na.last = TRUE)
ord_asc  <- order( ref$audit_score, na.last = TRUE)

.hyper_two_sided <- function(direction, ranked_genes, neg_set, top_n) {
  hit_g  <- ranked_genes[seq_len(top_n)]
  n_in   <- length(neg_set)
  hits   <- sum(hit_g %in% neg_set)
  expect <- top_n * n_in / n_atlas
  pval   <- if (direction == "depletion")
    phyper(hits, n_in, n_atlas - n_in, top_n, lower.tail = TRUE)
  else
    phyper(hits - 1, n_in, n_atlas - n_in, top_n, lower.tail = FALSE)
  data.table(direction = direction, top_n = top_n,
             hits = hits, expect = round(expect, 2),
             fold = round(hits / max(expect, 1e-9), 2),
             pval = formatC(pval, format = "g", digits = 2))
}
```

## C1. Top-N depletion (negative controls absent from the most-prioritised candidates)

```{r negctrl_top}
top_eval <- rbindlist(lapply(c(100, 500, 1000), function(N) {
  rbindlist(list(
    cbind(.hyper_two_sided("depletion",
            ref$gene_symbol[ord_desc], hk_set, N),
          neg_set = "housekeeping"),
    cbind(.hyper_two_sided("depletion",
            ref$gene_symbol[ord_desc], plh_set, N),
          neg_set = "plasma_high")))
}))
print(top_eval[, .(neg_set, top_n, hits, expect, fold, pval)])
```

## C2. Bottom-N enrichment (negative controls concentrated at the lowest-priority end)

This is the symmetric test: if the leakage gate is doing real
work, housekeeping and plasma_high genes should be **enriched**
in the bottom-ranked tail (the framework actively pushes them
down the list).

```{r negctrl_bot}
bot_eval <- rbindlist(lapply(c(100, 500, 1000), function(N) {
  rbindlist(list(
    cbind(.hyper_two_sided("enrichment",
            ref$gene_symbol[ord_asc], hk_set, N),
          neg_set = "housekeeping"),
    cbind(.hyper_two_sided("enrichment",
            ref$gene_symbol[ord_asc], plh_set, N),
          neg_set = "plasma_high")))
}))
print(bot_eval[, .(neg_set, top_n, hits, expect, fold, pval)])
```

**Interpretation.** Housekeeping genes show **strong, statistically
significant depletion** at the top of the audit-score ranking
(top-1000: 0 hits vs 3.07 expected, `p = 0.04`) and -- more
strikingly -- **strong enrichment at the bottom**: all 31
housekeeping genes land in the bottom 500 (`fold = 20.2`,
`p = 1.3e-41`). This is exactly what the leakage gate
(`audit_is_housekeeping -> 0.00`) is designed to do: drive
known artefactual genes to the score floor, regardless of how
strong their trajectory signal looks.

The plasma_high signal is weaker because their leakage
multiplier is `0.50` (not `0.00`), so they retain partial score
and are only mildly depleted.

Paired with the post-freeze anchor enrichment in Section B
(known PDAC biomarkers preferentially **at the top**), this
result demonstrates that the framework discriminates in **both
directions**: known signal goes up, known noise goes down. A
purely circular score would not produce both effects.

# Limitations of these analyses

- The post-freeze **evaluation set** (30 anchors, 14 T1) is
  **fixed at v0.3.0 freeze**. We do not use it as a training
  signal; we use it as a sanity-check reference that the
  framework was deliberately blinded to during construction.
  Anchor curation was completed before any audit-rule parameter
  was finalised, but anchor selection and framework design share
  domain expert input -- this is a **sanity check, not a fully
  blinded external validation**.
- The cutoff-sweep simulation in Section A1 perturbs the
  surface call without re-running the upstream DESeq2 LRT.
  Re-fitting from scratch at a different cutoff would not
  change the per-gene `rna_pattern_rho` itself (which is
  already stored), only which genes are surfaced. The
  sensitivity test is therefore exact, not an approximation.
- Section B compares pdactrace's audit ranking against
  *internal* simple rankings, not against external commercial
  tools (e.g., IPA, MetaCore). External-tool comparison is
  out of scope for an open-source R package vignette but would
  be a fair part of a manuscript benchmark.

```{r session_info}
sessionInfo()
```
