---
title: "5. Reproducibility from raw data"
author: "pdactrace"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Reproducibility from raw data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Goal

`pdactrace` ships a frozen 10,113-gene atlas plus the audit, query,
report, and projection API that consumes it. This vignette
documents **four reproducibility layers**, ordered from "no
external data needed" to "the entire FASTQ-to-atlas pipeline":

1. **Layer 1 — offline, bundled.** Use the prebuilt `data/*.rda`
   directly. No network, no rebuild.
2. **Layer 2 — user cohort.** Project a user-supplied count or
   intensity matrix through the same audit framework via
   `project_user_cohort()`.
3. **Layer 3 — bundled phase chain.** Re-derive the bundled atlas
   from the package's own `inst/extdata` plus the two large
   upstream phase CSVs (downloaded once via
   `download_phase_csvs()`).
4. **Layer 4 — full pipeline (FASTQ → counts → fits).**
   Out of scope for the package; this layer is the
   manuscript-monorepo's job. Pointer at the bottom.

# Layer 1 — Offline, bundled

```{r layer1, eval = FALSE}
library(pdactrace)

# All bundled atlas objects load lazily from `data/`:
data(pdactrace_reference)
data(pdactrace_protein_betas)
data(default_templates)
data(pdactrace_external_anchors)
data(meta_analysis)
data(atlas_metadata)
data(pdactrace_data_sources)

query_gene("LTBP1")             # uses pdactrace_reference internally
explain_score("LTBP1")          # same
list_data_sources(layer = "RNA")
atlas_provenance()
```

This is the canonical use case — most users never need anything
beyond Layer 1.

# Layer 2 — User cohort

If you have your own staged RNA-seq counts (and optionally a
matched tissue-protein intensity matrix), feed it into the same
framework:

```{r layer2, eval = FALSE}
data(toy_counts); data(toy_coldata); data(toy_protein)

res <- project_user_cohort(
  rna        = toy_counts,
  coldata    = toy_coldata,
  stage_col  = "stage",
  cohort_col = "cohort",
  protein    = toy_protein)

res$summary
head(res$audit[order(-audit_score)])
```

`project_user_cohort()` chains four steps internally:
`fit_stage_de()` → `classify_trajectory()` → `assemble_user_evidence()`
→ `compute_audit_score()`. Each can be called individually for
custom workflows; see `vignette("user_cohort_extension")`.

`SummarizedExperiment` input works identically — pass an `SE` whose
`assay()` is the count matrix and whose `colData()` carries the
stage / cohort columns (`fit_stage_de(se, stage_col, cohort_col)`).

# Layer 3 — Re-derive the bundled atlas

The full `data-raw/build_reference.R` chain reproduces
`data/pdactrace_reference.rda` from public phase CSVs. The package
already ships the **six small downstream tables** in
`inst/extdata/phase{2c,29,42,60,77,80}_*.csv.xz` (~52 KB
compressed total). The two large upstream fits — `phase33` (RNA)
and `phase34` (protein) — are too big to bundle and are pulled
once via `download_phase_csvs()`:

```{r layer3, eval = FALSE}
# 1. Pull the two large upstream CSVs (~3.5 MB total) into the user's
#    BiocFileCache. Subsequent calls reuse the cached copy.
paths <- download_phase_csvs("both")
paths
#> phase33  /home/.../BiocFileCache/.../phase33_deseq2_coef_12template.csv
#> phase34  /home/.../BiocFileCache/.../phase34_protein_pooled_12template.csv

# 2. The build_reference.R script's lookup chain
#    (inst/extdata -> $PDAC_BASE_DIR -> error) finds phase33/34 once
#    PDAC_BASE_DIR points to a local copy.  Easiest setup: clone the
#    manuscript-monorepo and point PDAC_BASE_DIR at it:
#       git clone https://github.com/jibeomko/PDAC_biomarker
#       export PDAC_BASE_DIR=$PWD/PDAC_biomarker
#       Rscript data-raw/build_reference.R
```

The build re-creates `data/pdactrace_reference.rda` byte-identical
to the bundled object, given the same upstream CSVs and the same
R / data.table versions.

The `build_meta_analysis.R`, `build_atlas_metadata.R`,
`build_audit_scores_v2.R`, etc. scripts under `data-raw/` extend
the same chain — all are deterministic given the bundled inputs +
the two cached CSVs.

For a one-off "fresh" rebuild without cloning the manuscript repo,
the snippet below works as long as both phase CSVs are reachable
on the local filesystem (regardless of where `PDAC_BASE_DIR`
points):

```{r layer3-direct, eval = FALSE}
# Manual rebuild path — pure R, no shell needed
paths <- download_phase_csvs("both")
target_dir <- file.path(
  Sys.getenv("PDAC_BASE_DIR", tempdir()),
  "analysis/manuscript/tissue_to_serum_biomarker/results")
dir.create(target_dir, recursive = TRUE, showWarnings = FALSE)
file.copy(paths, target_dir, overwrite = TRUE)

# Now `read_phase()` inside build_reference.R will resolve via the
# $PDAC_BASE_DIR fallback even if PDAC_BASE_DIR was just a tempdir.
Sys.setenv(PDAC_BASE_DIR = dirname(dirname(dirname(target_dir))))
source(system.file("..", "data-raw", "build_reference.R",
                    package = "pdactrace"))
```

# Layer 4 — From FASTQ

This layer — raw FASTQ → STAR/salmon counts → DESeq2 LRT → phase33
CSV — is **out of scope for `pdactrace`**. The required cohort-
specific Snakemake / Nextflow pipelines, ~110 numbered
`phase*` scripts, FragPipe configs, scVI / liana notebooks, and
~100 GB of intermediate artefacts live in the companion
manuscript-monorepo:

* GitHub: <https://github.com/jibeomko/PDAC_biomarker>
* Zenodo: [10.5281/zenodo.20067849](https://doi.org/10.5281/zenodo.20067849)

That repository carries the public-data accession manifests under
`analysis/metadata/`; `list_data_sources()` exposes the same
information programmatically inside `pdactrace`.

# Provenance summary

```{r prov, eval = FALSE}
ap <- atlas_provenance()
ap$package_doi
ap$manuscript_doi
ap$data_layers
list_data_sources()
```

Every gene's `provenance` field is renderable as a plain-English
evidence summary via [`format_provenance()`].
