---
title: "4. Apply pdactrace to your own multi-omics cohort"
author: "pdactrace v0.4.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Apply pdactrace to your own multi-omics cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(pdactrace)
  library(data.table)
})
set.seed(42)
```

# Goal

The bundled v0.3.0 reference atlas covers four PDAC RNA cohorts, two
PDAC tissue-protein cohorts, ten scRNA cohorts, and three serum-DIA
cohorts. v0.4.0 adds an additive, backwards-compatible API so a new
user can run the same frozen 3-axis + 2-gate audit framework on
their own multi-omics cohort, without retraining and without
modifying the bundled atlas. This vignette walks through the full
five-step workflow on a synthetic example so the framework's
input/output schema is concrete.

# Step 1 — Build a tiny synthetic counts matrix

```{r synth-rna}
n_genes <- 200
n_samples <- 24
sample_meta <- data.table(
  sample = paste0("S", 1:n_samples),
  cohort = rep(c("CohortA", "CohortB", "CohortC"), times = 8),
  stage  = rep(c("Normal", "Early", "Mid", "Late"), each = 6))

# Background expression
counts <- matrix(rnbinom(n_genes * n_samples, mu = 50, size = 4),
                  nrow = n_genes,
                  dimnames = list(paste0("GENE", 1:n_genes),
                                   sample_meta$sample))

# Inject 20 stage-progressive genes (Early_Burst_Up signature)
up_genes <- paste0("GENE", 1:20)
stage_lift <- c(Normal = 0, Early = 50, Mid = 70, Late = 80)
for (g in up_genes) {
  counts[g, ] <- counts[g, ] + stage_lift[as.character(sample_meta$stage)]
}
```

# Step 2 — Fit the stage-aware RNA model

```{r rna-fit, eval = requireNamespace("DESeq2", quietly = TRUE)}
rna_fit <- fit_stage_de(counts,
                          stage = sample_meta$stage,
                          cohort = sample_meta$cohort)
rna_pat <- classify_trajectory(rna_fit, rho_cutoff = 0.85)
head(rna_pat[!is.na(rna_pattern), .(gene_symbol, rna_pattern,
                                       rna_pattern_rho, lrt_padj)])
```

# Step 3 — Build a tiny synthetic protein matrix and fit limma

```{r prot-fit, eval = requireNamespace("limma", quietly = TRUE)}
prot_intensity <- matrix(
  rnorm(n_genes * n_samples, mean = 8, sd = 1),
  nrow = n_genes,
  dimnames = list(paste0("GENE", 1:n_genes), sample_meta$sample))
prot_lift <- c(Normal = 0, Early = 1.0, Mid = 1.5, Late = 1.8)
for (g in up_genes) {
  prot_intensity[g, ] <- prot_intensity[g, ] +
    prot_lift[as.character(sample_meta$stage)]
}

prot_fit <- fit_stage_de_protein(prot_intensity,
                                   stage = sample_meta$stage,
                                   cohort = sample_meta$cohort)
prot_pat <- classify_protein_trajectory(prot_fit, rho_cutoff = 0.85)
head(prot_pat[!is.na(prot_pattern), .(gene_symbol, prot_pattern,
                                         prot_pattern_rho)])
```

# Step 4 — Assemble the user evidence table

```{r assemble, eval = exists("rna_pat") && exists("prot_pat")}
user_ev <- assemble_user_evidence(
  rna_fit = rna_pat,
  prot_fit = prot_pat,
  signal_peptide = up_genes)
str(user_ev, max.level = 1)
```

# Step 5 — Run the audit framework on user evidence

```{r audit, eval = exists("user_ev")}
user_audit <- compute_audit_score(evidence = user_ev)
table(user_audit$audit_class)
head(user_audit[order(-audit_score)], 10)
```

The same `evaluate_anchor_enrichment()` interface accepts an
optional `anchors` table so a user can evaluate their own anchor
list:

```{r enrich, eval = exists("user_ev")}
user_anchors <- data.table(
  gene = up_genes[1:5],
  include_primary_eval = TRUE,
  include_secondary_eval = TRUE)
evaluate_anchor_enrichment(top_n = 20, tier = "secondary",
                             evidence = user_ev,
                             anchors = user_anchors)
```

# Step 6 — Single-patient trajectory alignment (v0.4.1)

`align_patient_profile()` is the single-sample analog of
`score_trajectory()`. It takes one patient's per-gene
log2(tumor / matched-normal) vector and reports how that profile
aligns with each of the atlas's frozen stage-effect axes
(`rna_beta_E`, `rna_beta_M`, `rna_beta_L`).

**This is an alignment readout, not a stage prediction.** No
supervised model has been fit — the atlas is frozen and the function
is deterministic given (input, atlas). Misalignment with all three
stage axes is itself a finding (it does not imply Normal).

```{r align_synthetic, eval = TRUE}
ref <- pdactrace:::.get_reference()
high_audit <- ref[!is.na(rna_beta_L) & audit_score >= 0.3][seq_len(300)]

# Synthetic patient that resembles a Late-stage profile
set.seed(7)
patient_rna <- setNames(
  high_audit$rna_beta_L + rnorm(nrow(high_audit), sd = 0.3),
  high_audit$gene_symbol)

aln <- align_patient_profile(patient_rna,
                               weight_by       = "audit_score",
                               min_audit_score = 0.3,
                               top_n_genes     = 200)
print(aln)
aln$rna
```

The output reports two complementary signals: the primary
audit-weighted Pearson rho between the patient and each stage axis
(with Fisher-z 95% CI), and a secondary audit-weighted vote share
showing what fraction of genes are closest to each stage's β. If
the two signals disagree, both are surfaced — they are not collapsed
into a single decision.

# Scope boundary

The PDAC reference atlas remains the default reference for
`compute_audit_score(evidence = NULL)` and `align_patient_profile()`;
the v0.4 release does not provide a `build_user_atlas()` function
or a layer plugin architecture. The frozen v0.3.0 scoring rule, the
audit labels, and the post-freeze evaluation discipline apply
identically whether the input is the bundled atlas or a
user-supplied evidence table — the framework is the same; only the
data source changes.

`align_patient_profile()` deliberately does **not** support
supervised stage prediction. A Random-Forest / GLM patient-level
classifier trained on TCGA / CPTAC stage labels is conceivable but
would break the package's central "no supervised fitting under
limited ground truth" discipline; it is therefore out of scope.
Limitations: rho is unstable below ~50 overlapping genes; the
protein layer reports only categorical concordance because the
atlas does not carry per-stage numeric protein effect sizes; the
audit-score floor restricts the gene dictionary to anchors with
multi-layer evidence — varying it changes which dictionary you align
against, not the patient. Disease-agnostic atlas building (Level 2)
is planned for v0.5.

```{r session_info}
sessionInfo()
```
