---
title: "`r params$gene` &mdash; pdactrace evidence report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: true
    toc_depth: 2
    df_print: kable
params:
  gene: "LTBP1"
---

```{css, echo=FALSE}
body {
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Helvetica,
               "Liberation Sans", sans-serif;
  font-size: 14px;
  line-height: 1.55;
  color: #1a1a1a;
  max-width: 920px;
  margin: 0 auto;
  padding: 28px 36px;
  background: #ffffff;
}
h1 { font-size: 22px; font-weight: 600; border-bottom: 1px solid #e2e2e2;
     padding-bottom: 8px; margin-top: 32px; }
h2 { font-size: 17px; font-weight: 600; margin-top: 26px; color: #2c2c2c; }
h3 { font-size: 14px; font-weight: 600; color: #4a4a4a; }
table { border-collapse: collapse; font-size: 12.5px; margin: 8px 0 14px; }
table th, table td { padding: 5px 10px; border-bottom: 1px solid #ececec;
                     text-align: left; }
table th { font-weight: 600; background: #fafafa; color: #333; }
code { background: #f5f5f5; padding: 1px 4px; border-radius: 2px;
       font-size: 12px; color: #c7254e; }
.tag { display: inline-block; padding: 2px 8px; border-radius: 3px;
       font-size: 12px; font-weight: 600; margin-right: 6px; }
.tag-high  { background: #1B5E20; color: white; }
.tag-supp  { background: #F57C00; color: white; }
.tag-pen   { background: #9E9E9E; color: white; }
.tag-excl  { background: #424242; color: white; }
.tag-low   { background: #BDBDBD; color: white; }
.summary-box { background: #fafbfc; border-left: 3px solid #0d47a1;
               padding: 10px 14px; margin: 12px 0; font-size: 13px; }
.muted { color: #6a6a6a; font-size: 12px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dev = "png",
  dpi = 150
)
suppressMessages({
  library(pdactrace)
  library(data.table)
})
gene <- params$gene
```

```{r resolve, include=FALSE}
ref <- pdactrace::pdactrace_reference
in_atlas <- gene %in% ref$gene_symbol
audit <- if (in_atlas) pdactrace::compute_audit_score(gene) else NULL
detailed <- if (in_atlas) tryCatch(pdactrace::query_gene_detailed(gene),
                                     error = function(e) NULL) else NULL
panel_row <- if (in_atlas) pdactrace::query_panel(gene) else NULL
```

`r if (!in_atlas) sprintf("> **%s** is not in the bundled pdactrace atlas (10,113 genes). Report cannot be generated.", gene)`

```{r assert-found, include=FALSE}
if (!in_atlas) knitr::knit_exit()
```

# Summary

```{r tag-html, results="asis"}
ac <- audit$audit_class[1]
tag_class <- switch(ac,
  high_confidence     = "tag-high",
  supported_uncertain = "tag-supp",
  penalized           = "tag-pen",
  excluded            = "tag-excl",
  "tag-low")
cat(sprintf(
  '<span class="tag %s">%s</span> <span class="muted">audit_score = %.3f</span>\n\n',
  tag_class, ac, audit$audit_score[1]))
```

<div class="summary-box">
**Atlas position:** `r gene` is one of `r format(nrow(ref), big.mark = ",")`
genes in the bundled atlas. The audit framework places it in the
*`r ac`* class with score
`r sprintf("%.3f", audit$audit_score[1])`
based on three evidence axes (multi-layer presence, biological
coherence, translational relevance) and two reliability gates
(housekeeping/leakage + cohort heterogeneity).
</div>

```{r summary-table}
key_cols <- c("rna_pattern", "prot_pattern", "translation_class",
              "cell_origin_top", "serum_detected", "rnaprot_concordant")
present <- intersect(key_cols, names(panel_row))
if (length(present) > 0) {
  out <- data.table::transpose(panel_row[, ..present], keep.names = "field")
  setnames(out, c("field", "value"))
  knitr::kable(out, format = "html", escape = TRUE,
               table.attr = 'style="width:auto;"')
}
```

# Audit components

```{r audit-table}
audit_cols <- c("evidence_strength", "biological_coherence",
                "translational_relevance", "leakage_gate",
                "heterogeneity_gate", "audit_score")
keep <- intersect(audit_cols, names(audit))
if (length(keep) > 0) {
  vals <- as.numeric(audit[1, ..keep])
  knitr::kable(
    data.frame(component = keep, value = sprintf("%.3f", vals)),
    format = "html", escape = TRUE,
    table.attr = 'style="width:auto;"')
}
```

# Evidence radar

```{r radar, fig.width=4.6, fig.height=3.6}
hex <- tryCatch(
  pdactrace::plot_gene_hexagon(gene, comparison = "high_confidence_mean"),
  error = function(e) NULL)
if (!is.null(hex)) print(hex) else cat("Hexagon not available for this gene.\n")
```

# Stage trajectory

```{r stage, fig.width=5.4, fig.height=3.0}
p_stage <- tryCatch(pdactrace::plot_stage_effect(gene),
                    error = function(e) NULL)
if (!is.null(p_stage)) print(p_stage) else cat("plot_stage_effect not available.\n")
```

# Per-cohort detail

```{r per-cohort, fig.width=5.4, fig.height=3.0}
p_cohort <- tryCatch(pdactrace::plot_per_cohort(gene),
                     error = function(e) NULL)
if (!is.null(p_cohort)) print(p_cohort) else cat("plot_per_cohort not available.\n")
```

# Per-stage detail

```{r per-stage, results="asis"}
if (!is.null(detailed) && !is.null(detailed$per_stage)) {
  cat('<div class="muted">RNA effect sizes per Normal / Early / Mid / Late stage.</div>\n\n')
  print(knitr::kable(detailed$per_stage, format = "html",
                     table.attr = 'style="width:auto;"'))
}
```

```{r per-cohort-table, results="asis"}
if (!is.null(detailed) && !is.null(detailed$per_cohort)) {
  cat("\n## Per-cohort breakdown\n\n")
  cat('<div class="muted">Cohort-level RNA / protein effect estimates.</div>\n\n')
  print(knitr::kable(detailed$per_cohort, format = "html",
                     table.attr = 'style="width:auto;"'))
}
```

# Filter trace

```{r filter-trace, fig.width=5.6, fig.height=3.0}
p_trace <- tryCatch(pdactrace::plot_filter_trace(gene),
                    error = function(e) NULL)
if (!is.null(p_trace)) print(p_trace) else cat("plot_filter_trace not available.\n")
```

# Provenance

```{r provenance, results="asis"}
meta <- tryCatch(pdactrace::list_atlas_metadata(),
                 error = function(e) NULL)
if (!is.null(meta)) {
  cat(sprintf("Atlas version `%s`, built on %s.\n\n",
              meta$version, meta$build_date))
}
```

<div class="muted">
Generated by pdactrace::report_gene(). The audit rule is frozen
(deterministic given input + atlas) and external anchors are used
for post-freeze evaluation only &mdash; this report does not represent
a supervised prediction. Misalignment with the high-confidence mean is
itself a useful diagnostic, not necessarily a failure mode.
</div>
