---
title: "pdactrace panel evidence report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    theme: null
    highlight: null
    toc: true
    toc_depth: 2
    df_print: kable
params:
  genes: !r c("LTBP1", "SERPINA1", "GAPDH")
---

```{css, echo=FALSE}
body {
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Helvetica,
               "Liberation Sans", sans-serif;
  font-size: 14px;
  line-height: 1.55;
  color: #1a1a1a;
  max-width: 980px;
  margin: 0 auto;
  padding: 28px 36px;
  background: #ffffff;
}
h1 { font-size: 22px; font-weight: 600; border-bottom: 1px solid #e2e2e2;
     padding-bottom: 8px; margin-top: 32px; }
h2 { font-size: 17px; font-weight: 600; margin-top: 26px; color: #2c2c2c; }
h3 { font-size: 14px; font-weight: 600; color: #4a4a4a; }
table { border-collapse: collapse; font-size: 12.5px; margin: 8px 0 14px; }
table th, table td { padding: 5px 10px; border-bottom: 1px solid #ececec;
                     text-align: left; }
table th { font-weight: 600; background: #fafafa; color: #333; }
code { background: #f5f5f5; padding: 1px 4px; border-radius: 2px;
       font-size: 12px; color: #c7254e; }
.tag { display: inline-block; padding: 2px 8px; border-radius: 3px;
       font-size: 12px; font-weight: 600; margin-right: 6px; }
.tag-high  { background: #1B5E20; color: white; }
.tag-supp  { background: #F57C00; color: white; }
.tag-pen   { background: #9E9E9E; color: white; }
.tag-excl  { background: #424242; color: white; }
.tag-low   { background: #BDBDBD; color: white; }
.summary-box { background: #fafbfc; border-left: 3px solid #0d47a1;
               padding: 10px 14px; margin: 12px 0; font-size: 13px; }
.gene-card { border: 1px solid #ececec; border-radius: 4px;
             padding: 14px 18px; margin: 14px 0; }
.muted { color: #6a6a6a; font-size: 12px; }
.explain-text { background: #fdfdfd; border-left: 2px solid #cfd8dc;
                padding: 8px 12px; font-size: 12.5px;
                white-space: pre-wrap; font-family: ui-monospace,
                "SF Mono", Menlo, monospace; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", dev = "png", dpi = 150)
suppressMessages({
  library(pdactrace)
  library(data.table)
})
genes <- params$genes
```

# Panel summary

`r length(genes)`-gene panel: **`r paste(genes, collapse = ", ")`**.

```{r compare-table}
cmp <- pdactrace::compare_candidates(genes)
knitr::kable(cmp, format = "html", escape = TRUE,
             table.attr = 'style="width:auto;"')
```

<div class="summary-box">
The audit framework places candidates into one of five classes:
`high_confidence`, `supported_uncertain`, `penalized`, `excluded`,
`low`. The score is deterministic given (gene, atlas) because the
v0.3.0 audit rule is frozen and external anchors are used for
post-freeze evaluation only. The `redundancy_with` column flags
genes that share both `rna_pattern` and `cell_origin_top` — useful
when picking a non-redundant panel.
</div>

# Panel-level evidence radar

```{r panel-hex, fig.width=5.4, fig.height=4.2}
in_atlas_genes <- genes[genes %in% pdactrace::pdactrace_reference$gene_symbol]
hex <- if (length(in_atlas_genes) > 0L) {
  tryCatch(pdactrace::plot_gene_hexagon(in_atlas_genes),
           error = function(e) NULL)
} else NULL
if (!is.null(hex)) print(hex) else
  cat("Hexagon not available for this panel.\n")
```

# Per-gene detail

```{r per-gene, results="asis"}
for (g in genes) {
  cat(sprintf('\n<div class="gene-card">\n<h2>%s</h2>\n', g))
  if (!g %in% pdactrace::pdactrace_reference$gene_symbol) {
    cat(sprintf('<p class="muted">%s is not in the bundled atlas.</p>\n',
                g))
    cat('</div>\n')
    next
  }
  ex <- pdactrace::explain_score(g, verbose = FALSE)
  tag_class <- switch(ex$audit_class,
    high_confidence     = "tag-high",
    supported_uncertain = "tag-supp",
    penalized           = "tag-pen",
    excluded            = "tag-excl",
    "tag-low")
  cat(sprintf(
    '<span class="tag %s">%s</span> <span class="muted">audit_score = %.3f</span>\n\n',
    tag_class, ex$audit_class, ex$audit_score))
  cat(sprintf('<div class="explain-text">%s</div>\n\n',
              gsub("<", "&lt;", ex$explanation, fixed = TRUE)))
  cat('</div>\n')
}
```

# Provenance

```{r provenance, results="asis"}
meta <- tryCatch(pdactrace::list_atlas_metadata(),
                 error = function(e) NULL)
if (!is.null(meta)) {
  cat(sprintf("Atlas version `%s`, built on %s.\n\n",
              meta$version, meta$build_date))
}
```

<div class="muted">
Generated by `pdactrace::report_gene(panel)`. The audit rule is
frozen (deterministic given input + atlas). External anchors are
used for post-freeze evaluation only — this report does not
represent a supervised prediction. Misalignment with the
`high_confidence` mean is itself a useful diagnostic, not necessarily
a failure mode.
</div>
